The demographic costs of sexually antagonistic selection in partially selfing populations

### 2021-06-04 doc/SA-Hermaphrodites-wDemography.tex
Pre-Submission Edits
This commit adds final edits to the main manuscript and online Supp.
Mat., includes new results for complementary sex-specific
dominance in the Supp. Mat., and removes author information for
double-blind peer review.

\author{Colin Olito$^{1,\ast}$ \\ 
Charlotte de Vries$^{2}$}
\date{\today}
\maketitle

\noindent{} 1. Department of Biology, Lund University, Lund 223 62, Sweden;

\noindent{} 2.  Department of Evolutionary Biology and Environmental Studies, University of Zurich, Zurich CH-8057, Switzerland;

\noindent{} $\ast$ Corresponding author; e-mail: colin.olito@gmail.com

@@

%\tableofcontents
%%%%%%%%%%%%%%%%%%%%
\newpage{}
\section*{Abstract -- currently 200/200 word max.}%\textcolor{green}{ I like this abstract best (CdV)} \textcolor{blue}{Agreed (CJO)}}

@@

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant (and often infinite) size, and therefore do not consider the population dynamical consequences of evolutionary change. In fact, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals (e.g., \citealt {johnston2013life, merot2020balancing}). To take into account the consequences of population dynamics under SA selection, a model is required that links changes in genotype distributions to population dynamics. The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few papers since then explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with intralocus sexual antagonism, and population dynamics but did not perform an analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

@@

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of partially-selfing simultaneous hermaphrodites. The derivation and some of the key results follow closely those presented in \citet{deVriesCaswell2019a} and \citet{deVriesCaswell2019b}. In fact, the model reduces to the stage-structured model of \citet{deVriesCaswell2019a} under obligate outcrossing. A full derivation of the model and analyses is presented in the \hl{Online Supplementary Material}, and all computer code necessary to reproduce the results are available at \url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}.

Simultaneous hermaphrodites express both male and female sex functions (i.e., can transmit genes to the next generation via both sperm/pollen and eggs/ovules), and have the potential to reproduce by a combination of self- and outcross-fertilizations. Maternal outcrossing involves receiving male gametes from another individual in the population, while paternal outcrossing involves exporting male gametes to another individual and fertilizing their ovules. Self-fertilization is achieved when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female function, we denote matrices or vectors relating to male sex function with a prime.

@@

The population projection matrix $\tilde{\mbf{A}}$ is constructed from four sets of matrices representing the demographic and genetic processes: The matrices $\mbf{U}^{S}_{i}$ and $\mbf{U}^{X}_{i}$ contain transition and survival probabilities for each genotype, produced by selfing and outcrossing respectively. The matrices $\mbf{F}_{i}$ and $\mbf{F}^{\prime}_{i}$ contain the genotype $\times$ stage specific contribution via female and male sex-functions of genotype $i$ to the female and male gamete pools, respectively, and therefore to zygotes in the next generation. We assume that whether individuals were produced through selfing or outcrossing does not affect their fecundity or mating success, that is, we assume  $\mbf{F}^X_{i}=\mbf{F}^S_{i}=\mbf{F}_{i}$ and $\mbf{F}^{\prime X}_{i}=\mbf{F}^{\prime S}_{i}=\mbf{F}^{\prime}_{i}$. Deviations from this assumption are straightforward to incorporate but beyond the scope of this paper.

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Mating and offspring production under partial selfing}
 Outcrossing in our model proceeds similarly to a two-sex model of reproduction \citep{deVriesCaswell2019b}. That is, each individual's genotype determines both the number of ovules produced and the paternal "mating success", broadly defined. For hermaphroditic flowering plants, for example, paternal mating success could reflect pollen production, export efficiency, pollen-tube germination and growth rates, among other things \citep{LloydWebb1986, WangBarrett2020, Harder2016}. Note, however, that by modeling paternal relative mating success rather than pollen/sperm production, we implicitly assume female demographic dominance (i.e., production/transport of male gametes does not limit ovule fertilization and hence population growth).

@@

Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen, for reasons of analytic tractability. In the \hl{Online Supplement}, we present a more general model of genotype-specific self-fertilization rates (after \citealt{JordanConnallon2014}).

%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Population projection}
Using the component matrices described above (the survival matrices, $\mathbf{U}^X_i$, $\mathbf{U}^S_i$, the fertility and mating success matrices, $\mathbf{F}_i$, $\mathbf{F}^\prime_i$, and the parent-offspring matrices $\mbf{H}^S(\tilde{\mbf{n}})$ and $\mbf{H}^X$), we construct the population projection matrix $\tilde{\mbf{A}}[\tilde{\mbf{n}}]$ using the vec-permutation approach of \citet{CaswellEtAl2018}, see \hl{Online Supplementary Material} for the step by step construction of the model. The matrix that projects the eco-evolutionary dynamics is:
\begin{linenomath*}
\begin{equation} \label{eq:Atilde}
	\tilde{\mbf{A}}[\tilde{\mbf{n}}] = 
			\underbrace{\left(
			\begin{array}{c|c}
				\mcal{U}^S & \mbf{0} \\ \hline
				\mbf{0} & \mcal{U}^X \\
			\end{array} \right)}_{\tilde{\mbf{U}}} + 
			\underbrace{\left(
			\begin{array}{c|c}
				\mcal{F}^S(\tilde{\mbf{p}}) C(1 - \delta) & \mcal{F}^S(\tilde{\mbf{p}}) C(1 - \delta) \\ \hline
				\mcal{F}^X(\tilde{\mbf{p}}) (1 - C) & \mcal{F}^X(\tilde{\mbf{p}}) (1 - C)\\
			\end{array} \right)}_{\tilde{\mbf{F}}},
\end{equation}
\end{linenomath*}
where $C$ denotes the proportion of each individual's ovules that are self-fertilized (the remaining $1 - C$ are outcrossed), and  $\delta$ represents the proportion of self-fertilized zygotes that fail to develop due to inbreeding depression during early development \citep{Charlesworth1987}.

The blocks of the component matrices in Eq(\ref{eq:Atilde}) correspond to production of offspring by self-fertilization and outcrossing ($\mcal{F}^S$ and $\mcal{F}^X$ in $\tilde{\mbf{F}}$), and survival of extant individuals produced by selfing or outcrossing ($\mcal{U}^S$ and $\mcal{U}^X$ in $\tilde{\mbf{U}}$). The survival matrices for individuals produced through selfing and outcrossing are,
\begin{linenomath*}
\begin{eqnarray} 
	\mcal{U}^S  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{S}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{S}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{S}_{aa} \\
			\end{array} \right),\label{eq:BlkUS}\\
				\mcal{U}^X  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{X}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{X}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{X}_{aa} \\
			\end{array} \right).\label{eq:BlkUX}
\end{eqnarray}
\end{linenomath*}
Since individuals cannot change their genotype once they are born, the survival matrices are block diagonal.  Similarly, we construct fertility matrices for individuals produced through selfing and outcrossing, 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFS}
	\mcal{F}^S = 
			\left(
			\begin{array}{ccc}
				\mbf{F}_{AA} & \frac{1}{4} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{2} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{4} \mbf{F}_{Aa} & \mbf{F}_{aa}\\
			\end{array} \right), 
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFX}
	\mcal{F}^X = 
			\left(
			\begin{array}{ccc}
				q^{\prime}_{A} \mbf{F}_{AA} & \frac{1}{2} q^{\prime}_{A} \mbf{F}_{Aa} & 0 \\ 
				q^{\prime}_{a} \mbf{F}_{AA} & \frac{1}{2} \mbf{F}_{Aa} & q^{\prime}_{A} \mbf{F}_{aa}  \\ 
				0 & \frac{1}{2} q^{\prime}_{a} \mbf{F}_{Aa} & q^{\prime}_{a} \mbf{F}_{aa}  \\
			\end{array} \right),
\end{equation}
\end{linenomath*}
\noindent where $q^{\prime}_A$ and $q^{\prime}_a$ are given by Eq(\ref{eq:maleGametePool}). Further details of the derivations of $\mcal{U}^S$, $\mcal{U}^X$ ,  $\mcal{F}^S$, and $\mcal{F}^X$ are provided in the \hl{Online Supplement}.

@@

Combining all the component matrices yields the overall eco-evolutionary projection matrix shown in \hl{Appendix A}.

@@

$s_f,\,s_m$ & Selection coefficients through female and male reproductive function & \\
$h_f,\,h_m$ & Dominance coefficients through female and male reproductive function & \\
$\sigma_j,\, \sigma_a$ & Survival rates for juvenile and adult stages & \\
$\gamma$ & Transition rate from juvenile to adult stages \\
$C$ & The population rate of self-fertilization & \\
$\delta,\, \delta_j,\,\delta_a,\,\delta_{\gamma}$ & Inbreeding depression terms for ovule viability, survival, and transition rates respectively & \\
$f_i,\,f^{\prime}_i$ & Adult fertilities through male and female sex functions & \\

@@

We now construct and analyze a genotype $\times$ stage-classified model for a hypothetical species with intralocus sexual conflict via the two sex functions. For the sake of simplicity, we assume our hypothetical species has a basic life cycle with only two life stages: juveniles and adults (i.e., $\omega = 2$), and that only adults are reproductively active. Suppose that there is a genetic trade-off between the sex-functions at a single diallelic locus such that allele $A$ is beneficial for female fertility but detrimental for male reproductive success (e.g., pollen production), and that allele $a$ has the reverse effect. Following convention, we parameterize the fertility component of fitness for each genotype through each sex function, $w_{i}$ and $w^{\prime}_{i}$, to be bounded by $[0,1]$, with dominance and selection coefficients $h_f, s_f$ and $h_m, s_m$ determining the decrease in fertility through each sex function relative to the most fit genotype ($AA$ has highest female fertility, $aa$ the highest male fertility; see Table \ref{tab:Fitness}).

The SA locus does not affect survival and transition rates. However, the survival matrices can be used to model the fitness effects of inbreeding depression at later stages of development by allowing different stage-specific survival and transition rates for individuals produced by self-fertilization vs.~outcrossing. By contrast, the parameter $\delta$ only affects inbreeding depression through viability of selfed ovules. With this in mind, we define survival matrices for individuals produced by selfing and outcrossing as follows:
\begin{linenomath*}
\begin{equation} \label{eq:US}
	\mbf{U}^S = \left(
					\begin{array}{cc}
						\sigma_j (1 - \delta_j) \big(1 - \gamma (1 - \delta_{\gamma}) \big) & 0 \\
						\sigma_j (1 - \delta_j) \gamma (1 - \delta_{\gamma})      & \sigma_a (1 - \delta_a)
					\end{array}
				\right) \\
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation}\label{eq:UX}
	\mbf{U}^X = \left(
					\begin{array}{cc}
						\sigma_j(1 - \gamma) & 0 \\
						\sigma_j \gamma      & \sigma_a
					\end{array}
				\right)
\end{equation}
\end{linenomath*}

@@

Unless stated otherwise, we use the following parameter values for the demographic rates in the model: $\sigma_j = \sigma_a = 0.6$ and $\gamma = 0.05$. These chosen values are similar to those used in \citet{deVriesCaswell2019b}, facilitating comparison between models, and to correspond to a life-history in which individuals spend multiple timesteps in the juvenile phase prior to maturing into reproductively active adults, but are otherwise arbitrary. Our parameters of interest include fertility, $f$, the inbreeding depression parameters, $\delta$, $\delta_j$, $\delta_a$, $\delta_{\gamma}$, and the selection parameters $h_f$, $s_f$, $h_m$, and $s_m$, which are given different values for each analyses as described in the figure captions.

@@

We identify conditions where SA polymorphism is 'protected' by evaluating the stability of populations initially fixed for either SA allele to invasion by the other (i.e., we assessed stability at the boundary equilibrium genotype frequencies of $\hat{\mbf{p}}_{AA} = 1$ and $\hat{\mbf{p}}_{aa} = 1$; \citealt{Levene1953, Prout1968, deVriesCaswell2019b}). The formal conditions for a protected polymorphism are determined by linearizing the model in the vicinity of the boundary equilibria ($\hat{\mbf{p}}$), and evaluating the magnitude of the largest eigenvalue of the Jacobian matrix of the linearization. A full derivation of the Jacobian and details of the invasion analysis are provided in the \hl{Online Supplementary Material}, and the relevant leading eigenvalues are presented in \hl{Appendix B}.

@@

% Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. 
Because single-locus selection coefficients are generally weak (e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. We present scenarios of equal dominance in the main text (i.e., $h_f = h_m = h$). Specifically, we examine scenarios of ({\itshape i}) additive SA fitness effects ($h = 1/2$), which are commonly observed for small-effect alleles or quantitative traits \citep{AgrawalWhitlock2011}; and ({\itshape ii}) dominance reversals, where the deleterious fitness effect of each SA allele is partially recessive in each sex ($h = 1/4$), which are predicted under fitness landscape models provided the population is not too far from the phenotypic optimum \citep{Manna2011, ConnallonClark2014}. We briefly explore the consequences of sex-specific dominance, which may evolve under asymmetric SA selection \citep{SpencerPriest2016} in \hl{Appendix X}).

@@

We begin with an illustration of demographically viable polymorphic parameter space in the absence of inbreeding depression in Figure \ref{fig:extThresholds} (i.e., $\delta = \delta_i = 0$). The red areas in Figure \ref{fig:extThresholds} indicate areas where the population goes extinct (population growth rate is less than one) for different values of the fertility parameter.  Lower female fertility values correspond to a larger demographically inviable area. The funnel between the thick black lines indicates the area where the sexually antagostic alleles coexist. Above the funnel, the female-benefit and male-detriment allele fixates; below the funnel, the male-benefit and female-detriment allele fixates. Much of the region where the male-benefit and female-detriment allele fixates is demographically inviable due to the demographic consequences of reduced female fitness in this region.  The SA polymorphisms that remain viable at lower fertility values correspond to regions where the female-deleterious allele is predicted to segregate at low frequencies, or in regions with weak selection through both sex functions, an asymmetry that reflects the assumption of female demographic dominance.

@@

In populations with high fertility (larger $f$), the proportion of demographically viable polymorphic parameter space converges on the predictions for total SA polymorphic space in population genetic models (fig.~\ref{fig:polySpace}). In obligately outcrossing populations (including dioecious/gonochoristic populations; where $C = 0$), lower fertility can result in a significant reduction of demographically viable polymorphic parameter space. However, the effect is weaker in populations with intermediate selfing rates (compare $C=0$ vs.~$C > 0$) because self-fertilization protects a fraction, $C$, of the ovules of individuals homozygous for the female-beneficial allele from exposure to the male-beneficial allele. In other words, self-fertilization reduces the opportunity for selection favouring the female-deleterious allele, thereby reducing the equilibrium load on female fecundity and allowing partially selfing populations to remain viable under selection intensities that would cause extinction in an outcrossing population. The combination of protection from reduced female fertility and reduced total polymorphic parameter space caused by selfing results in populations with intermediate selfing rates having the greatest proportion of demographically viable parameter space at medium and low fertilities (fig.~\ref{fig:polySpace}, med.~and low.~fertility values).

@@

@@

As an illustrative example of how our model can be used to explore whether demographic rates observed in natural populations appear likely to support balanced SA polymorphisms, we parameterized the model using empirically estimated demographic rates and fitness data for natural populations of the hermaphroditic flowering plant {\itshape Mimulus guttatus} (Scrophulariaceae; now known as {\itshape Erythranthe guttata}). {\itshape M.~guttatus} is an herbaceous, self-compatible wildflower native to western North America that exhibits remarkable among-population variation in numerous life-history and reproductive traits including selfing rates, inbreeding depression, floral morphology, and annual-to-perennial life-history \citep[e.g.,][]{RitlandGanders1987, Ritland1990, Willis1993, Willis1999a, Willis1999b, WuWillis2008}. Moreover, detailed demographic studies have been conducted on multiple populations of {\itshape M.~guttatus}, with demographic data available on the public demographic database COMPADRE \citep{CompadreDB2020}. Below, we briefly outline how we parameterized our model using the available data; full details are provided in \hl{Appendix C}.

>>>>

%\author{Colin Olito$^{1,\ast}$ \\ 
%Charlotte de Vries$^{2}$}
\date{\today}
\maketitle

%\noindent{} 1. Department of Biology, Lund University, Lund 223 62, Sweden;

%\noindent{} 2.  Department of Evolutionary Biology and Environmental Studies, University of Zurich, Zurich CH-8057, Switzerland;

%\noindent{} $\ast$ Corresponding author; e-mail: colin.olito@gmail.com

@@

%\tableofcontents
%%%%%%%%%%%%%%%%%%%%
\newpage{}
\section*{Abstract}%\textcolor{green}{ I like this abstract best (CdV)} \textcolor{blue}{Agreed (CJO)}}

@@

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant (and often infinite) size, and therefore do not consider the population dynamical consequences of evolutionary change. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals (e.g., \citealt {johnston2013life, merot2020balancing}). To take into account the consequences of population dynamics under SA selection, a model is required that links changes in genotype distributions to population dynamics. The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few papers since then have explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with intralocus sexual antagonism, and population dynamics but did not perform an analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

@@

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of partially-selfing simultaneous hermaphrodites. The derivation and some of the key results follow closely those presented in \citet{deVriesCaswell2019a} and \citet{deVriesCaswell2019b}. In fact, the model reduces to the stage-structured model of \citet{deVriesCaswell2019a} under obligate outcrossing. A full derivation of the model and analyses is presented in the Online Supplementary Material, and all computer code necessary to reproduce the results are available at \url{GitHub Link omitted for anonymity}.%\url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}.

Simultaneous hermaphrodites can transmit genes to the next generation via both sperm/pollen and eggs/ovules, and have the potential to reproduce by a combination of self- and outcross-fertilizations. Maternal outcrossing involves receiving male gametes from another individual in the population, while paternal outcrossing involves exporting male gametes to another individual and fertilizing their ovules. Self-fertilization is achieved when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female function, we denote matrices or vectors relating to the male sex function with a prime.

@@

The population projection matrix $\tilde{\mbf{A}}$ is constructed from four sets of matrices representing the demographic and genetic processes: The matrices $\mbf{U}^{S}_{i}$ and $\mbf{U}^{X}_{i}$ contain transition and survival probabilities for each genotype, produced by selfing and outcrossing respectively. The matrices $\mbf{F}_{i}$ and $\mbf{F}^{\prime}_{i}$ contain the genotype $\times$ stage specific contribution of genotype $i$ to the female and male gamete pools, respectively, and therefore to zygotes in the next generation. We assume that whether individuals were produced through selfing or outcrossing does not affect their fecundity or mating success, that is, we assume  $\mbf{F}^X_{i}=\mbf{F}^S_{i}=\mbf{F}_{i}$ and $\mbf{F}^{\prime X}_{i}=\mbf{F}^{\prime S}_{i}=\mbf{F}^{\prime}_{i}$. Deviations from this assumption are straightforward to incorporate but beyond the scope of this paper.

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Mating and offspring production under partial selfing}
 Outcrossing in our model proceeds similarly to a two-sex model of reproduction \citep{deVriesCaswell2019b}. That is, each individual's genotype determines both the number of ovules produced and the paternal mating success, broadly defined. For hermaphroditic flowering plants, for example, paternal mating success could reflect pollen production, export efficiency, pollen-tube germination and growth rates, among other things \citep{LloydWebb1986, WangBarrett2020, Harder2016}. Note, however, that by modeling paternal relative mating success rather than pollen/sperm production, we implicitly assume female demographic dominance (i.e., production/transport of male gametes does not limit ovule fertilization and hence population growth), a point we return to in the Discussion.

@@

Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen, for reasons of analytic tractability. In the Online Supplement, we present a more general model of genotype-specific self-fertilization rates (after \citealt{JordanConnallon2014}).

%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Population projection}
Using the component matrices described above (the survival matrices, $\mathbf{U}^X_i$, $\mathbf{U}^S_i$, the fertility and mating success matrices, $\mathbf{F}_i$, $\mathbf{F}^\prime_i$, and the parent-offspring matrices $\mbf{H}^S(\tilde{\mbf{n}})$ and $\mbf{H}^X$), we construct the population projection matrix $\tilde{\mbf{A}}[\tilde{\mbf{n}}]$ using the vec-permutation approach of \citet{CaswellEtAl2018}, see Online Supplementary Material for the step by step construction of the model. The matrix that projects the eco-evolutionary dynamics is:
\begin{linenomath*}
\begin{equation} \label{eq:Atilde}
	\tilde{\mbf{A}}[\tilde{\mbf{n}}] = 
			\underbrace{\left(
			\begin{array}{c|c}
				\mcal{U}^S & \mbf{0} \\ \hline
				\mbf{0} & \mcal{U}^X \\
			\end{array} \right)}_{\tilde{\mbf{U}}} + 
			\underbrace{\left(
			\begin{array}{c|c}
				\mcal{F}^S(\tilde{\mbf{p}}) C(1 - \delta) & \mcal{F}^S(\tilde{\mbf{p}}) C(1 - \delta) \\ \hline
				\mcal{F}^X(\tilde{\mbf{p}}) (1 - C) & \mcal{F}^X(\tilde{\mbf{p}}) (1 - C)\\
			\end{array} \right)}_{\tilde{\mbf{F}}},
\end{equation}
\end{linenomath*}
where $C$ denotes the proportion of each individual's ovules that are self-fertilized (the remaining $1 - C$ are outcrossed), and  $\delta$ represents the proportion of self-fertilized zygotes that fail to develop due to inbreeding depression during early development \citep{Charlesworth1987}.

The blocks of the component matrices in Eq(\ref{eq:Atilde}) correspond to production of offspring by self-fertilization and outcrossing ($\mcal{F}^S$ and $\mcal{F}^X$ in $\tilde{\mbf{F}}$), and survival of extant individuals produced by selfing or outcrossing ($\mcal{U}^S$ and $\mcal{U}^X$ in $\tilde{\mbf{U}}$). The survival matrices for individuals produced through selfing and outcrossing are,
\begin{linenomath*}
\begin{eqnarray} 
	\mcal{U}^S  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{S}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{S}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{S}_{aa} \\
			\end{array} \right),\label{eq:BlkUS}\\
				\mcal{U}^X  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{X}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{X}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{X}_{aa} \\
			\end{array} \right).\label{eq:BlkUX}
\end{eqnarray}
\end{linenomath*}
Since individuals cannot change their genotype once they are born, the survival matrices are block diagonal.  Similarly, we construct fertility matrices for individuals produced through selfing and outcrossing, 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFS}
	\mcal{F}^S = 
			\left(
			\begin{array}{ccc}
				\mbf{F}_{AA} & \frac{1}{4} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{2} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{4} \mbf{F}_{Aa} & \mbf{F}_{aa}\\
			\end{array} \right), 
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFX}
	\mcal{F}^X = 
			\left(
			\begin{array}{ccc}
				q^{\prime}_{A} \mbf{F}_{AA} & \frac{1}{2} q^{\prime}_{A} \mbf{F}_{Aa} & 0 \\ 
				q^{\prime}_{a} \mbf{F}_{AA} & \frac{1}{2} \mbf{F}_{Aa} & q^{\prime}_{A} \mbf{F}_{aa}  \\ 
				0 & \frac{1}{2} q^{\prime}_{a} \mbf{F}_{Aa} & q^{\prime}_{a} \mbf{F}_{aa}  \\
			\end{array} \right),
\end{equation}
\end{linenomath*}
\noindent where $q^{\prime}_A$ and $q^{\prime}_a$ are given by Eq(\ref{eq:maleGametePool}).

@@

Combining all the component matrices yields the overall eco-evolutionary projection matrix shown in Appendix A.

@@

$s_f,\,s_m$ & Selection coefficients through female and male reproductive function & \\
$h_f,\,h_m$ & Dominance coefficients through female and male reproductive function & \\
$\sigma_j,\, \sigma_a$ & Survival rates for juvenile and adult stages & \\
$\gamma$ & Transition rate from juvenile to adult stages \\
$C$ & The population rate of self-fertilization & \\
$\delta,\, \delta_j,\,\delta_a,\,\delta_{\gamma}$ & Inbreeding depression terms for ovule viability, survival, and transition rates respectively & \\
$f_i,\,f^{\prime}_i$ & Adult fertilities through female and male sex functions respectively & \\

@@

We now construct and analyze a genotype $\times$ stage-classified model for a hypothetical species with intralocus sexual conflict via the two sex functions. For the sake of simplicity, we assume our hypothetical species has a life cycle with only two stages: juveniles and adults (i.e., $\omega = 2$), and that only adults are reproductively active. Suppose that there is a genetic trade-off between the sex-functions at a single diallelic locus such that allele $A$ is beneficial for female fertility but detrimental for male reproductive success (e.g., pollen production), and that allele $a$ has the reverse effect. Following convention, we parameterize the fertility component of fitness for each genotype through each sex function, $w_{i}$ and $w^{\prime}_{i}$, to be bounded by $[0,1]$, with dominance and selection coefficients $h_f, s_f$ and $h_m, s_m$ determining the decrease in fertility through each sex function relative to the most fit genotype ($AA$ has highest female fertility, $aa$ the highest male fertility; see Table \ref{tab:Fitness}).

The SA locus does not affect survival or transition rates. However, the survival matrices can be used to model the fitness effects of inbreeding depression at later stages of development by allowing different stage-specific survival and transition rates for individuals produced by self-fertilization vs.~outcrossing. By contrast, the parameter $\delta$ only affects inbreeding depression through viability of selfed ovules. With this in mind, we define survival matrices for individuals produced by selfing and outcrossing as follows:
\begin{linenomath*}
\begin{equation} \label{eq:US}
	\mbf{U}^S = \left(
					\begin{array}{cc}
						\sigma_j (1 - \delta_j) \big(1 - \gamma (1 - \delta_{\gamma}) \big) & 0 \\
						\sigma_j (1 - \delta_j) \gamma (1 - \delta_{\gamma})      & \sigma_a (1 - \delta_a)
					\end{array}
				\right) \\
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation}\label{eq:UX}
	\mbf{U}^X = \left(
					\begin{array}{cc}
						\sigma_j(1 - \gamma) & 0 \\
						\sigma_j \gamma      & \sigma_a
					\end{array}
				\right)
\end{equation}
\end{linenomath*}

@@

Unless stated otherwise, we use the following parameter values for the demographic rates in the model: $\sigma_j = \sigma_a = 0.6$ and $\gamma = 0.05$. These chosen values are similar to those used in \citet{deVriesCaswell2019b}, facilitating comparison between models, and correspond to a life-history in which individuals spend multiple timesteps in the juvenile phase prior to maturing into reproductively active adults, but are otherwise arbitrary. Our parameters of interest include fertility, $f$, the inbreeding depression parameters, $\delta$, $\delta_j$, $\delta_a$, $\delta_{\gamma}$, and the selection parameters $h_f$, $s_f$, $h_m$, and $s_m$, which are given different values for each analyses as described in the figure captions.

@@

We identify conditions where SA polymorphism is 'protected' by evaluating the stability of populations initially fixed for either SA allele to invasion by the other (i.e., we assessed stability at the boundary equilibrium genotype frequencies of $\hat{\mbf{p}}_{AA} = 1$ and $\hat{\mbf{p}}_{aa} = 1$; \citealt{Levene1953, Prout1968, deVriesCaswell2019b}). The formal conditions for a protected polymorphism are determined by linearizing the model in the vicinity of the boundary equilibria ($\hat{\mbf{p}}$), and evaluating the magnitude of the largest eigenvalue of the Jacobian matrix of the linearization. A full derivation of the Jacobian and details of the invasion analysis are provided in the Online Supplementary Material, and the relevant leading eigenvalues are presented in Appendix B.

@@

% Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. 
Because single-locus selection coefficients are generally weak (e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. We present scenarios of equal dominance in the main text (i.e., $h_f = h_m = h$). Specifically, we examine scenarios of ({\itshape i}) additive SA fitness effects ($h = 1/2$), which are commonly observed for small-effect alleles or quantitative traits \citep{AgrawalWhitlock2011}; and ({\itshape ii}) dominance reversals, where the deleterious fitness effect of each SA allele is partially recessive in each sex ($h = 1/4$), which are predicted under fitness landscape models provided the population is not too far from the phenotypic optimum \citep{Manna2011, ConnallonClark2014}. We briefly explore the consequences of sex-specific dominance in the Online Supplementary Material.

@@

We begin with an illustration of demographically viable polymorphic parameter space in the absence of inbreeding depression in Figure \ref{fig:extThresholds} (i.e., $\delta = \delta_i = 0$). The red regions in Figure \ref{fig:extThresholds} indicate areas where the population goes extinct (population growth rate is less than one) for different values of the fertility parameter. Lower female fertility values correspond to a larger demographically inviable area. The funnel between the thick black lines indicates the area where the sexually antagostic alleles coexist. Above the funnel, the female-benefit and male-detriment allele goes to fixation; below the funnel, the male-benefit and female-detriment allele fixates. Much of the region where the male-benefit and female-detriment allele fixates is demographically inviable due to the demographic consequences of reduced female fitness in this region. The SA polymorphisms that remain viable at lower fertility values correspond to regions where the female-deleterious allele is predicted to segregate at low frequencies, or in regions with weak selection through both sex functions, an asymmetry that reflects the assumption of female demographic dominance.

@@

@@

In populations with high fertility (larger $f$), the proportion of demographically viable polymorphic parameter space converges on the predictions for total SA polymorphic space in population genetic models (fig.~\ref{fig:polySpace}). In obligately outcrossing populations (including dioecious/gonochoristic populations; where $C = 0$), lower fertility can result in a significant reduction of demographically viable polymorphic parameter space. The effect is weaker in populations with intermediate selfing rates (compare $C=0$ vs.~$C > 0$) because self-fertilization generates a greater proportion of offspring that are homozygous for the female-beneficial allele relative to heterozygotes. In other words, self-fertilization reduces the opportunity for selection favouring the female-deleterious allele, thereby reducing the equilibrium load on female fecundity and allowing partially selfing populations to remain viable under selection intensities that would cause extinction in an outcrossing population. The combination of protection from reduced female fertility and reduced total polymorphic parameter space caused by selfing results in populations with intermediate selfing rates having the greatest proportion of demographically viable parameter space at medium and low fertilities (fig.~\ref{fig:polySpace}, med.~and low.~fertility values).

@@

As an illustrative example of how our model can be used to explore whether demographic rates observed in natural populations appear likely to support balanced SA polymorphisms, we parameterized the model using empirically estimated demographic rates and fitness data for natural populations of the hermaphroditic flowering plant {\itshape Mimulus guttatus} (Scrophulariaceae; now known as {\itshape Erythranthe guttata}). {\itshape M.~guttatus} is an herbaceous, self-compatible wildflower native to western North America that exhibits remarkable among-population variation in numerous life-history and reproductive traits including selfing rates, inbreeding depression, floral morphology, and annual-to-perennial life-history \citep[e.g.,][]{RitlandGanders1987, Ritland1990, Willis1993, Willis1999a, Willis1999b, WuWillis2008}. Moreover, detailed demographic studies have been conducted on multiple populations of {\itshape M.~guttatus}, with demographic data available on the public demographic database COMPADRE \citep{CompadreDB2020}. Below, we briefly outline how we parameterized our model using the available data; full details are provided in Appendix C.
